### 2021-04-04 paper.Rmd
proofreading and commenting the first draft #3

Nonlinear dimension reduction techniques such as t-SNE have become very popular in recent years, primarily for the ability to capture cluster structure in a succinct visual summary. However, it is only a summary, and it likely involves substantial warping of the original data space. Using the tour along with these techniques can illuminate the nature of the warped space, and reveal other structure lost in the dimension reduction.  Figure \ref{fig:tsne-tour} illustrates the difference in what can be learned from a tour in comparison with nonlinear dimension reduction using t-SNE. The 10-dimensional data comes from @Rauber2009-kk. The t-SNE view (plot A) shows six clearly separated clusters, all spherical with different sizes. The tour shows that the clusters look very little like this in the full data space. The clusters are at various distances apart and very different sizes. In the four projections from the tour this can be seen. The two green clusters are large and spherical, and far from the brown clusters. The brown clusters have one larger one, and three smaller, very close to each other. All of these are elliptical, which means that they have very little variability, actually no variability if you watch the full tour, in some of the 10 dimensions. This gives some deeper perspective to what is learned from t-SNE, and illustrates what the t-SNE dimension reduction has done: it has found small gaps between points and expanded these gaps to yield the representation. It should be noted, though, that the t-SNE view clearly reveals the presence of six distinct clusters. With this information is the invitation to look closer at the data in the tour, to see, that yes, indeed, there are three, tiny, tiny clusters very close to each other.

@@

The rest of the review is structured as follows: in section \@ref(notation) we define the notation and components of a tour display. By its nature a tour is most effective to analyst when combined with interactivity; in section \@ref(interaction) we review the components of user interfaces for manipulating tour views. The implementation of tour paths in statistical software is reviewed in section \@ref(software), where we discuss the history and state of the art packages for computing them. Section \@ref(applications) shows the diverse applications of the tour in the natural sciences, machine learning and applied statistics. Finally, in section \@ref(discussion) we point to future research directions for tours.

@@

Below introduces the notations used for describing $d$-dimensional projections in a $p$-dimensional space. Let $X_{n \times p}$ be the data whose projections are of interest. A projection basis, $A_{p \times d}$, characterises the direction from which the data are projected and needs to satisfy an orthonormality condition, which requires each column to be a unit vector and to be perpendicular to each other. This condition gives the $p$-dimensional basis space a geometrical structure of a Stiefel manifold. With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. Figure \ref{fig:notation-huber} shows an example of the 2D iris data (sepal measures only) in a huber's plot, along with a 1D projection in the histogram. This projection is created based on projecting the data onto the solid line in the huber plot and it separates group setosa from versicolor and virginica. (This plot is created with package PPtreeViz, whose reference is @Lee2005-gl)

\begin{figure*}
   \centering
         \includegraphics[width=\textwidth]{figures/notation-huber.pdf}
\caption{Huber's plot (left) of the 2D iris data (sepal length and width only). The solid ellipse represents how well the projection, onto different directions, separates the three species and the dashed circle is a reference guide. The histogram (right) is the 1D projection of the data onto the direction outlined as the solid line in the huber's plot.}
\label{fig:notation-huber}
\end{figure*}

@@

To define whether a projection is interesting or not in the guided tour, an index function is used. An index function defines a surjection function from the projection $Y$ in $\mathbb{R}^{n \times d}$ to a scalar in $\mathbb{R}$. For the high-dimensional data, an interesting projection is the one that deviates from normality, as discussed in @Diaconis1984-mv and @Huber1985-zq [expand more here]. Numerous index functions have been developed since the initial index proposed by @Friedman1974-ck: <!-- entropy-based index? --> holes and central mass index [@Cook2008-nl] computes the deviation of the projected data from a normal distribution; @Posse1995-ih proposed a chi-square index for detecting the spiral structure in the data;  kurtosis and skewness indices are raised in @Loperfido2020-xj and @Loperfido2018-ga to detect outliers in financial time series; and most recently, scagnostic indices [@Laa2020-zj] provides eight measures for summarising information between pairs of variables in scatter plot matrices.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/notation-target-gen.pdf}
\caption{A comparison of guided and grand tour path in the PCA reduced space. The start point of tour paths is smaller and ligher than the end point. Guided tour optimises the index value iteratively and finishes the search quickly while grand tour wanders in the parameter space to look at possible interesting projections.}
\label{fig:notation-target-gen}
\end{figure}

@@

The smooth animation between the selected target frame is generated via a step-wise rotation in the high-dimensional space. Different methods to derive this rotation were described in @Buja2005-cx, and we can distinguish two types of interpolation. When interpolating between frames we rotate the starting basis onto the exact basis representation selected as a basis, and @Buja2005-cx describes three different methods to achieve this (based on decompositions of orthogonal matrices, Givens decompositions, or Householder decompositions). Alternatively, when interpolating between planes the rotation is constructed towards any basis that represents the plane corresponding to the target frame, such that it follows the shortest path, i.e. the geodesic, avoiding any in-plane rotation. The geodesic interpolation is derived using singular value decomposition and is implemented in the `tourr` package.

@@

The classic tour displays work with projections, and interactive systems like GGobi also provide the option of adding sectioning information via linked brushing. Sectioning means we select points that fall in a section of the full parameter space, for example for the purpose of highlighting them in a tour display. This often reveals complementary information, and combining sections and projections can for example provide insights into the dimensionality of a data structure [@Furnas1994-dz].

@@

The slice tour can reveal concave or non-linear structures obscured in projections, as well as small structures hidden near the center of a distribution. An example showing snapshots of the slice tour of points distributed on the surface of geometric shapes is shown in Fig. \ref{fig:tour-slice}. As with projections, an index function may be used to select target planes showing interesting slices of the data, and we can define a section pursuit guided tour [@Laa2020-ni].

@@

Transformations of the data are common in the preprocessing, examples would be scaling, sphering, or a logarithmic transformation for skewed distributions. However, we may also want to transform the data after projecting onto lower dimensions, to correct for unwanted effects of the projections. One example is the piling effect that is observed when projecting a high-dimensional distribution [@Diaconis1984-mv]: projected points are approximately Gaussian in most views, and increasingly concentrating near the center as dimensionality increases. The sage display [@Laa2020-uv] proposes a solution to this issue, via a radial transformation of the projected data. This non-linear transformation is defined in each point of the projection plane and is sensitive to the overall scale of the data as well as the original dimensionality $p$. In addition, tuning parameters can be used to obtain a more (or less) aggressive redistribution of the points. This is illustrated in Fig. \ref{fig:pollen} showing the application of the sage transformation to the classical pollen dataset [@Coleman1986-so]. The projection without any rescaling looks similar to what is found when rescaling with default options (left), and we can use either of the two tuning parameters to better resolve the distribution near the center and reveal the hidden structure (middle and right).

\begin{figure}
\centering
\includegraphics[width=0.3\textwidth]{figures/pollen_sage-088.png}
\includegraphics[width=0.3\textwidth]{figures/pollen_sage_R1-088.png}
\includegraphics[width=0.3\textwidth]{figures/pollen_sage_gam20-088.png}
\caption{Snapshots of the same projection from the sage tour on the pollen data, illustrating the different tuning options. Left: using the default settins for the transformation results in a similar view as found in projections without the sage transformation. Middle: using the tuning parameter $R$ for better resolution near the center reveals the word "EUREKA". Right: using the $\gamma$ tuning parameter can also reveal the structure.}
\label{fig:pollen}
\end{figure}

@@

Manual tours [@cook_manual_1997; @spyrison_spinifex_2020] offer a means to interactively control the contribution of a single variable on the projection plane. This is particularly useful for exploring a projection once a feature of interest has been identified. Manual tours can then be employed to explore local area of, or test the structure of the feature, with respect to a selected variable.

@@

\begin{figure*}[!ht]
   \centering
         \includegraphics[width=0.32\textwidth]{figures/penguins_manual_bl1_r.pdf}
         \includegraphics[width=0.32\textwidth]{figures/penguins_manual_bl3_r.pdf}
         \includegraphics[width=0.32\textwidth]{figures/penguins_manual_bl4_r.pdf}
\caption{Four projections from a manual tour where variable "bl" is being rotated out of the projection (orange line). When this variable is removed the two light green clusters merge, which informs us that "bl" is an important variable for distinguishing between these two groups.}
\label{fig:manual-tour}
\end{figure*}

@@

@@

\begin{figure*}[!ht]
   \centering
         \includegraphics[width=0.45\textwidth]{figures/ggobi_book/brush-query.pdf}
         \includegraphics[width=0.45\textwidth]{figures/ggobi_book/brush-response.pdf}
\caption{When a user brushes data on the left panel, points that fall inside the region are highlighted in orange. The view on the right responds by higlighting the corresponding points.}
\label{fig:brushing}
\end{figure*}

>>>>

Nonlinear dimension reduction techniques such as t-SNE have become very popular in recent years, primarily for the ability to capture cluster structure in a succinct visual summary. However, it is only a summary, and it likely involves substantial warping of the original data space. Using the tour along with these techniques can illuminate the nature of the warped space, and reveal other structure lost in the dimension reduction.  Figure \ref{fig:tsne-tour} illustrates the difference in what can be learned from a tour in comparison with nonlinear dimension reduction using t-SNE. The 10-dimensional data comes from @Rauber2009-kk. The t-SNE view (plot A) shows six clearly separated clusters, all spherical with different sizes. The tour shows that the clusters look very little like this [EW: not sure what "like this" referes to] in the full data space. The clusters are at various distances apart and very different sizes. In the four projections from the tour this can be seen. The two green clusters are large and spherical, and far from the brown clusters. The brown clusters have one larger one, and three smaller, very close to each other. All of these are elliptical, which means that they have very little variability, actually no variability if you watch the full tour, in some of the 10 dimensions. This gives some deeper perspective to what is learned from t-SNE, and illustrates what the t-SNE dimension reduction has done: it has found small gaps between points and expanded these gaps to yield the representation. It should be noted, though, that the t-SNE view clearly reveals the presence of six distinct clusters. With this information is the invitation to look closer at the data in the tour, to see, that yes, indeed, there are three, tiny, tiny clusters very close to each other.

@@

The rest of the review is structured as follows: in section \@ref(notation) we [EW: all of a sudden, from third person to first person?] define the notation and components of a tour display. By its nature a tour is most effective to analyst when combined with interactivity; in section \@ref(interaction) we review the components of user interfaces for manipulating tour views. The implementation of tour paths in statistical software is reviewed in section \@ref(software), where we discuss the history and state of the art packages for computing them. Section \@ref(applications) shows the diverse applications of the tour in the natural sciences, machine learning and applied statistics. Finally, in section \@ref(discussion) we point to future research directions for tours.

@@

Below introduces the notations used for describing $d$-dimensional projections in a $p$-dimensional space. Let $X_{n \times p}$ be the data whose projections are of interest. A projection basis, $A_{p \times d}$, characterizes the direction from which the data are projected and needs to satisfy an orthonormality condition, which requires each column to be a unit vector and to be perpendicular to each other. This condition gives the $p$-dimensional basis space a geometrical structure of a Stiefel manifold. With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. Figure \ref{fig:notation-huber} shows an example of the 2D iris data (sepal measures only) in a huber's plot, along with a 1D projection in the histogram. This projection is created based on projecting the data onto the solid line in the huber's plot and it separates group setosa from versicolor and virginica. (This plot is created with package PPtreeViz, whose reference is @Lee2005-gl)

\begin{figure*}
   \centering
         \includegraphics[width=\textwidth]{figures/notation-huber.pdf}
\caption{Huber's plot (left) of the 2D iris data (sepal length and width only). The solid ellipse represents how well the projection, onto different directions, separating the three species and the dashed circle is a reference guide. The histogram (right) is the 1D projection of the data onto the direction outlined as the solid line in the huber's plot.}
\label{fig:notation-huber}
\end{figure*}

@@

To define whether a projection is interesting or not in the guided tour [EW: should we add a new section on tour types before this section?], an index function is used. An index function defines a surjection function from the projection $Y$ in $\mathbb{R}^{n \times d}$ to a scalar in $\mathbb{R}$. For the high-dimensional data, an interesting projection is the one that deviates from normality, as discussed in @Diaconis1984-mv and @Huber1985-zq [expand more here]. Numerous index functions have been developed since the initial index proposed by @Friedman1974-ck: <!-- entropy-based index? --> holes and central mass index [@Cook2008-nl] computes the deviation of the projected data from a normal distribution; @Posse1995-ih proposed a chi-square index for detecting the spiral structure in the data;  kurtosis and skewness indices are raised in @Loperfido2020-xj and @Loperfido2018-ga to detect outliers in financial time series; and most recently, scagnostic indices [@Laa2020-zj] provide eight measures for summarizing information between pairs of variables in SPLOM.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/notation-target-gen.pdf}
\caption{A comparison of guided and grand tour path in the PCA reduced space. The start point of tour paths is smaller and lighter than the end point. Guided tour optimises the index value iteratively and finishes the search quickly while grand tour wanders in the parameter space to look at possible interesting projections.}
\label{fig:notation-target-gen}
\end{figure}

@@

The smooth animation between the selected target frames is generated via a step-wise rotation in the high-dimensional space. Different methods to derive this rotation were described in @Buja2005-cx, and we can distinguish two types of interpolation [EW: ??]. When interpolating between frames we rotate the starting basis onto the exact basis representation selected as a basis, and @Buja2005-cx describes three different methods to achieve this (based on decompositions of orthogonal matrices, Givens decompositions, or Householder decompositions). Alternatively, when interpolating between planes the rotation is constructed towards any basis that represents the plane corresponding to the target frame, such that it follows the shortest path, i.e. the geodesic, avoiding any in-plane rotation. The geodesic interpolation is derived using singular value decomposition and is implemented in the `tourr` package. [EW: this paragraph perhaps needs some rework.]

@@

Interactive systems like GGobi provide the option of adding sectioning information via linked brushing. Sectioning means we select points that fall in a section of the full parameter space, for example for the purpose of highlighting them in a tour display. This often reveals complementary information, and combining sections and projections can for example provide insights into the dimensionality of a data structure [@Furnas1994-dz].

@@

The slice tour can reveal concave or non-linear structures obscured in projections, as well as small structures hidden near the center of a distribution. Fig. \ref{fig:tour-slice} shows snapshots of the slice tour of points distributed on the surface of geometric shapes. Along with projections, an index function can be used to select target planes to show interesting slices of the data, defined as a section pursuit guided tour [@Laa2020-ni].

@@

Transformations of the data are common in the preprocessing [EW: "preprocessing" means b/f projecting?], such as scaling, sphering, or a logarithmic transformation for skewed distributions. However, we may also want to transform the data after projecting onto lower dimensions, to correct for unwanted effects of the projections. One example is the piling effect that is observed when projecting a high-dimensional distribution [@Diaconis1984-mv]: projected points are approximately Gaussian in most views, and increasingly concentrating near the center as dimensionality increases. The sage display [@Laa2020-uv] proposes a solution to this issue, via a radial transformation of the projected data. This non-linear transformation is defined in each point of the projection plane and is sensitive to the overall scale of the data as well as the original dimensionality $p$. In addition, tuning parameters can be used to obtain a more (or less) aggressive redistribution of the points. This is illustrated in Fig. \ref{fig:pollen} showing the application of the sage transformation to the classical pollen dataset [@Coleman1986-so]. The projection without any rescaling looks similar to what is found when rescaling with default options (left), and we can use either of the two tuning parameters to better resolve the distribution near the center and reveal the hidden structure (middle and right).

\begin{figure}
\centering
\includegraphics[width=0.3\textwidth]{figures/pollen_sage-088.png}
\includegraphics[width=0.3\textwidth]{figures/pollen_sage_R1-088.png}
\includegraphics[width=0.3\textwidth]{figures/pollen_sage_gam20-088.png}
\caption{Snapshots of the same projection from the sage tour on the pollen data, illustrating the different tuning options. Left: using the default settings for the transformation results in a similar view as found in projections without the sage transformation. Middle: using the tuning parameter $R$ for better resolution near the center reveals the word "EUREKA". Right: using the $\gamma$ tuning parameter can also reveal the structure.}
\label{fig:pollen}
\end{figure}

@@

Manual tours [@cook_manual_1997; @spyrison_spinifex_2020] offer a means to interactively control the contribution of a single variable on the projection plane. This is particularly useful for exploring a projection once a feature of interest has been identified. Manual tours can then be employed to explore local area of interest, or test the structure of the feature, with respect to a selected variable.

@@

\begin{figure*}[!ht]
   \centering
         \includegraphics[width=0.32\textwidth]{figures/penguins_manual_bl1_r.pdf}
         \includegraphics[width=0.32\textwidth]{figures/penguins_manual_bl3_r.pdf}
         \includegraphics[width=0.32\textwidth]{figures/penguins_manual_bl4_r.pdf}
\caption{Three projections from a manual tour where variable "bl" is being rotated out of the projection (orange line). When this variable is removed the two light green clusters merge, which informs us that "bl" is an important variable for distinguishing between these two groups.}
\label{fig:manual-tour}
\end{figure*}

@@

[EW: seems unrelated to the rest of the article; alternatively briefly explain "brushing" in s2.5?]

@@

\begin{figure*}[!ht]
   \centering
         \includegraphics[width=0.45\textwidth]{figures/ggobi_book/brush-query.pdf}
         \includegraphics[width=0.45\textwidth]{figures/ggobi_book/brush-response.pdf}
\caption{When a user brushes data on the left panel, points that fall inside the region are highlighted in orange. The view on the right responds by highlighting the corresponding points.}
\label{fig:brushing}
\end{figure*}
